Natural selection in the Health and Retirement Study

### 2023-09-29 hrs-selection.qmd
Add intergenerational correlation & bootstraps

The HRS sample focuses on cohorts born between 1920 and 1960, but contains some younger and older participants. I include only male participants born before 1965 and female participants born before 1970, which guarantees that most will have completed their fertility by 2010.

The resulting sample contains `r n_white` genotyped white participants and `r n_black` genotyped black participants. Genotyping took place in 2006, 2008 and subsequent years. I discard obsolete PGS for which there is a newer, more accurate score targeting the same phenotype. I also discard PGS for number of children ever born (but keep scores for age at first birth). This leaves a total of `r length(pgs)` scores.

PGS are rescaled to zero mean and unit variance within each ethnic group, so coefficient sizes are not directly comparable between ethnic groups, but are measured in standard deviations of the within-ethnic-group score. In all regressions using PGS, I control for ten within-ethnicity principal components of the DNA array data.

@@

```{r}
#| label: fig-rlrs-edu
#| fig-cap: |
#|   Scatterplot of PGS regression coefficients on RLRS and years of
#|   education. Controls include 10 principal components of genetic array
#|   data. Dashed lines show linear regressions.
#| fig-subcap:
#|   - ""
#|   - ""
#| layout-nrow: 2
#| column: page

@@

I can also examine natural selection in the previous generation, by regressing PGS on respondents' number of living siblings in 2010. (Data for dead siblings has too many missing values to use.) I reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent. Parents of no siblings cannot be included, so coefficient sizes are not comparable across the generations. Appendix @fig-siblings plots coefficients on number of siblings versus coefficients on years of education. Correlations are insignificant for both ethnic groups, although note that standard errors are large (whites: correlation `r cors_sibs_edu$white[1]`, bootstrapped 95% C.I. `r ci_boot_sibs$white[1]` to `r ci_boot_sibs$white[2]`; blacks: correlation `r cors_sibs_edu$black[1]`, bootstrapped 95% C.I. `r ci_boot_sibs$black[1]` to `r ci_boot_sibs$black[2]`).

>>>>

The HRS sample focuses on cohorts born between 1920 and 1960, but contains some younger and older participants. I include only male participants born before 1965 and female participants born before 1970, which guarantees that most will have completed their fertility by 2010. The resulting sample contains `r n_white` genotyped white participants and `r n_black` genotyped black participants.

Genotyping took place in 2006, 2008 and subsequent years. I discard obsolete PGS for which there is a newer, more accurate score targeting the same phenotype. I also discard PGS for number of children ever born (but keep scores for age at first birth). This leaves a total of `r length(pgs)` scores. PGS are rescaled to zero mean and unit variance within each ethnic group, so coefficient sizes are not directly comparable between ethnic groups, but are measured in standard deviations of the within-ethnic-group score. In all regressions using PGS, I control for ten within-ethnicity principal components of the DNA array data.

@@

```{r}
#| label: fig-rlrs-edu
#| fig-cap: |
#|   Scatterplot of PGS regression coefficients on RLRS against 
#|   coefficients on years of education. Each dot is one polygenic score.
#|   Controls include 10 principal components of genetic array
#|   data. Dashed lines are fitted from linear regressions.
#| fig-subcap:
#|   - ""
#|   - ""
#| layout-nrow: 2
#| column: page

@@

```{r}
#| label: stat-bootstrap-gen-cor
#| cache: true

n_reps_gen_cor <- 199

f_resp <- map(pgs, \(x) reformulate(c(x, pcs), response = "raevbrn"))
f_sibs <- map(pgs, \(x) reformulate(c(x, pcs), response = "r10sibs"))

estimate_gen_cor <- function (w, data) {
  data$w <- w # avoid w getting lost by map and lm
  m_resp <- map(f_resp, \(x) lm(x, data, weights = w))
  m_sibs <- map(f_sibs, \(x) lm(x, data, weights = w/r10sibs))

  coef_resp <- map2_dbl(m_resp, pgs, \(x, y) coef(x)[[y]])
  coef_sibs <- map2_dbl(m_sibs, pgs, \(x, y) coef(x)[[y]])
  c(gen_cor = cor(coef_resp, coef_sibs))
}

rand_boot <- rand |> 
    subset(ethnicity == "white") |> 
    as.svrepdesign(type = "bootstrap", replicates = n_reps_gen_cor)

gen_cor <- withReplicates(rand_boot, theta = estimate_gen_cor)
ci_gen_cor <- confint(gen_cor)
```

I can also examine natural selection in the previous generation, by regressing PGS on respondents' number of living siblings in 2010. (Data for dead siblings has too many missing values to use.) I reweight respondents by the reciprocal of their number of siblings, to account for parents of many siblings being more likely to be a parent of a respondent. Parents of no siblings cannot be included, so coefficient sizes are not comparable across the generations. Appendix @fig-siblings plots coefficients on number of siblings versus coefficients on years of education. Correlations are insignificant for both ethnic groups, with large standard errors (whites: correlation `r cors_sibs_edu$white[1]`, bootstrapped 95% C.I. `r ci_boot_sibs$white[1]` to `r ci_boot_sibs$white[2]`; blacks: correlation `r cors_sibs_edu$black[1]`, bootstrapped 95% C.I. `r ci_boot_sibs$black[1]` to `r ci_boot_sibs$black[2]`). However, there is a positive and significant correlation across generations, i.e. between PGS coefficients on number of siblings and number of children, of `r gen_cor`, bootstrapped 95% C.I. `r ci_gen_cor[1,1]` to `r ci_gen_cor[1,2]`.
